---
title: "What Drives National Happiness? A Data-Driven Analysis"
format: html
editor: visual
---

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(pander)
library(broom)
```

## Introduction

Understanding what contributes to happiness on a national scale has become an increasingly important topic in global policy and social science. As governments seek to improve the happiness of its citizens, data-driven analysis can provide valuable insights that go beyond traditional economic metrics. The World Happiness Report serves as a comprehensive resource by combining subjective life evaluations with key socio-economic indicators.

In this project, we analyze data from the 2023 World Happiness Report (WHR23) to explore how national characteristics relate to self-reported life satisfaction. Using a combination of descriptive statistics, multiple linear regression, Principal Component Analysis (PCA), and k-means clustering, we aim to uncover both individual predictors of happiness and broader patterns across countries.

# [Exploration and Analysis]{style="font-size:0.9em"}

## [Dataset Overview]{style="font-size:0.8em"}

Each observation in the WHR23 dataset corresponds to a country, and the main outcome variable is the average **Ladder Score**, which measures citizens’ self-reported life evaluation on a scale from 0 (worst possible life) to 10 (best possible life).

The dataset includes the following predictors:

## [Dataset Overview]{style="font-size:0.8em"}

-   **Logged GDP per capita**: Natural log of GDP adjusted for purchasing power parity.
-   **Social support**: Country average of a binary response on whether people have someone to rely on in times of trouble.
-   **Healthy life expectancy**: Expected years of life in good health.
-   **Freedom to make life choices**: Country average of a binary response on whether people are satisfied with their freedom to make life choices.
-   **Generosity**: Per-capita donations adjusted for income level.
-   **Perceptions of corruption**: average of the responses of two binary questions:
    -   Is corruption widespread throughout the government or not? ($Q_1$)
    -   Is corruption widespread within businesses or not? ($Q_2$)
    -   Corruption Index = $\frac{\text{Mean}(Q_1) + \text{Mean}(Q_2)}{2}$

These variables are used to explain cross-country variation in happiness levels.

## [1. Descriptive Analysis]{style="font-size:0.8em"}

#### 1.1 Summary Statistics

To get a general sense of global happiness in 2023, we first take a look at some summary statistics of Ladder scores across countries.

```{r, echo=FALSE}
whr23 <- read.csv("data/WHR23.csv")
```

::: {.model-table style="border:2px solid #ccc; padding:0.1em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
whr23_summary <- whr23 %>%
  summarise(
    Mean = mean(Ladder.score, na.rm = TRUE),
    Median = median(Ladder.score, na.rm = TRUE),
    Max = max(Ladder.score, na.rm = TRUE),
    Min = min(Ladder.score, na.rm = TRUE),
    Std = sd(Ladder.score, na.rm = TRUE),
    Q1 = quantile(Ladder.score, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(Ladder.score, probs = 0.75, na.rm = TRUE)
  ) %>% round(2)

pander(whr23_summary, caption = "Summary Statistics of Ladder Scores")
```
:::

**Interpretation:** The average happiness score across all countries in 2023 is approximately **5.54**, with the highest being **7.80 (Finland)** and the lowest being **1.86 (Afghanistan)**. The median score is slightly higher at **5.68**, indicating that more than half of the countries have Ladder Scores below the mean. The interquartile range shows that **50%** of countries have scores between **4.72** (Q1) and **6.33** (Q3).

## [1. Descriptive Analysis]{style="font-size:0.8em"}

#### 1.2 Distribution of Ladder Scores

We then create a histogram of Ladder Scores in order to visualize their distribution.

::::: columns
::: {.column width="50%"}
```{r, echo=FALSE, fig.height=6}
ggplot(whr23, aes(x = Ladder.score)) +
  geom_histogram(bins = 25, fill = "lightblue", color = "white") +
  labs(
    title = "Distribution of Ladder Scores",
    x = "Ladder Score",
    y = "Number of Countries"
  ) +
  theme_minimal(base_size = 20)
```
:::

::: {.column width="50%"}
**Interpretation:** The distribution of happiness scores is approximately bell-shaped but slightly left-skewed. This supports our earlier observation that more countries have Ladder Scores below the global average, indicating a concentration of nations with moderate to low levels of reported well-being.
:::
:::::

We then take a look at the top 10 and bottom 10 happiest countries based on their average Ladder Scores.

## [1. Descriptive Analysis]{style="font-size:0.8em"}

#### 1.3 Top 10 Happiest Countries

```{r, echo=FALSE, fig.align='center'}
top10 <- whr23[order(whr23$Ladder.score, decreasing = TRUE),
               c("Country.name", "Ladder.score")]
top10 <- top10[1:10,]

top10 %>%
  ggplot(aes(x = Ladder.score, 
             y = factor(Country.name, levels = rev(Country.name)))) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_text(aes(label = round(Ladder.score, 2)), 
            hjust = 1.5, size = 3.5) +
  labs(x = "Ladder Score", 
       y = "Country Name", 
       title = "Top 10 Countries by Ladder Score") +
  theme_minimal(base_size = 14)
```

**Interpretation:** Finland ranks as the happiest country in 2023, followed closely by Denmark and Iceland. The top 10 is largely composed of European nations, all of which are developed countries, reflecting regional and developmental trends in overall well-being.

## [1. Descriptive Analysis]{style="font-size:0.8em"}

#### 1.4 Bottom 10 Happiest Countries

```{r, echo=FALSE, fig.align='center'}
bottom10 <- whr23[order(whr23$Ladder.score, decreasing = FALSE),
               c("Country.name", "Ladder.score")]
bottom10 <- bottom10[1:10,]

bottom10 %>%
  ggplot(aes(x = Ladder.score, 
             y = factor(Country.name, levels = rev(Country.name)))) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_text(aes(label = round(Ladder.score, 2)), 
            hjust = 1.5, size = 3.5) +
  labs(x = "Ladder Score", 
       y = "Country Name", 
       title = "Bottom 10 Countries by Ladder Score") +
  theme_minimal(base_size = 14)
```

**Interpretation:** The bottom 10 countries in terms of happiness scores are led by Afghanistan, followed by Lebanon and Sierra Leone. Most of these countries face significant challenges such as political instability, economic hardship, or conflict. The list is predominantly composed of developing countries, particularly from sub-Saharan Africa and Middle East, highlighting a stark contrast in well-being compared to more developed regions.

## [2. Bivariate Relationships with Ladder Score]{style="font-size:0.8em"}

#### 2.1 Happiness vs. GDP per Capita

Whether higher income levels are associated with greater well-being? We examine the relationship between GDP per capita and the Ladder Score:

```{r, echo=FALSE, message=FALSE, fig.align='center'}
ggplot(whr23, aes(x = Logged.GDP.per.capita, y = Ladder.score)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Happiness vs. GDP per Capita",
    x = "Logged GDP per Capita",
    y = "Ladder Score"
  ) +
  theme_minimal(base_size = 14)
```

## [2. Bivariate Relationships with Ladder Score]{style="font-size:0.8em"}

#### 2.2 Happiness vs. Social Support

Since social connections may influence well-being, we use the plot below to explore the relationship between social support and happiness.

```{r, echo=FALSE, message=FALSE, fig.align='center'}
ggplot(whr23, aes(x = Social.support, y = Ladder.score)) +
  geom_point(color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Happiness vs. Social Support",
    x = "Social Support",
    y = "Ladder Score"
  ) +
  theme_minimal(base_size = 14)
```

## [2. Bivariate Relationships with Ladder Score]{style="font-size:0.8em"}

#### 2.3 Happiness vs. Freedom to Make Life Choices

Freedom to make life choices is another key aspect of personal well-being. We analyze the relationship between perceived freedom and the Ladder Score across countries.

```{r, echo=FALSE, message=FALSE, fig.align='center'}
ggplot(whr23, aes(x = Freedom.to.make.life.choices, y = Ladder.score)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Happiness vs. Freedom to Make Life Choices",
    x = "Freedom to Make Life Choices",
    y = "Ladder Score"
  ) +
  theme_minimal(base_size = 14)
```

## [2. Bivariate Relationships with Ladder Score]{style="font-size:0.8em"}

#### 2.4 Happiness vs. Perceptions of Corruption

Perceived corruption can weaken social trust and governance, reducing well-being. Let's examines its relationship with Ladder Scores across countries.

```{r, echo=FALSE, message=FALSE, fig.align='center'}
ggplot(whr23, aes(x = Perceptions.of.corruption, y = Ladder.score)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Happiness vs. Perception of corruption",
    x = "Perceptions of Corruption",
    y = "Ladder Score"
  ) +
  theme_minimal(base_size = 14)
```

## [3. Multiple Linear Regression Model]{style="font-size:0.8em"}

#### 3.1 Full Model

To assess the combined impact of multiple factors on happiness, we fit a multiple linear regression model using the six predictors from the dataset.

:::: panel-tabset
### Coefficients Table

::: {.model-table style="border:2px solid #ccc; padding:0.1em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
whr23 <- na.omit(whr23)
model <- lm(Ladder.score ~ Logged.GDP.per.capita +
              Social.support + Healthy.life.expectancy +
              Freedom.to.make.life.choices +
              Generosity + Perceptions.of.corruption,
            data = whr23)

tidy(model) %>% pander()
```
:::

### Interpretation

-   $R^2$ = 0.828

The multiple linear regression model explains approximately **82.8%** of the variation in happiness scores. Statistically significant predictors include **GDP per capita**, **social support**, **freedom to make life choices**, and **perceptions of corruption**, all of which are associated with happiness in the expected directions. Notably, **social support** has the strongest positive effect on the Ladder Score among all predictors in the model.
::::

## [3. Multiple Linear Regression Model]{style="font-size:0.8em"}

#### 3.2 Reduced Model

Since some variables have high p-values, we simplify the model by sequentially removing the predictor with the highest p-value until only statistically significant variables remain, using a 5% significance level. We begin by removing Generosity, which has a p-value of 0.66.

::: {.model-table style="border:2px solid #ccc; padding:0.2em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
reduced_model_1 <- lm(Ladder.score ~ Logged.GDP.per.capita +
                        Social.support + 
                        Healthy.life.expectancy +
                        Freedom.to.make.life.choices +
                        Perceptions.of.corruption,
                      data = whr23)

tidy(reduced_model_1) %>% pander()
```
:::

-   $R^2$ = 0.8278

## [3. Multiple Linear Regression Model]{style="font-size:0.8em"}

#### 3.2 Reduced Model

Next, we remove the variable Healthy.life.expectancy with a p-value of 0.16.

::: {.model-table style="border:2px solid #ccc; padding:0.2em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
reduced_model_2 <- lm(Ladder.score ~ Logged.GDP.per.capita +
                        Social.support + 
                        Freedom.to.make.life.choices +
                        Perceptions.of.corruption,
                      data = whr23)

tidy(reduced_model_2) %>% pander()
```
:::

-   $R^2$ = 0.8251

All remaining variables are statistically significant, so we consider this our final model.

```{r, echo=FALSE}
final_model <- reduced_model_2
```

## [3. Multiple Linear Regression Model]{style="font-size:0.8em"}

#### 3.3 ANOVA Test

To evaluate whether the additional predictors in the full model significantly improve the model fit, we perform an ANOVA test.

::: {.model-table style="border:2px solid #ccc; padding:0em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
anova_result <- anova(final_model, model)
knitr::kable(anova_result, caption = "ANOVA Test Comparing Reduced and Full Models")
```
:::

-   **Interpretation:**
    -   The ANOVA test yields a p-value of 0.3439, indicating that adding **Healthy life expectancy** and **Generosity** does not significantly improve the model.

In context, this suggests that life expectancy and donations per capita are less central to explaining cross-country differences in happiness compared to other factors.

## [4. Model Diagnostics]{style="font-size:0.8em"}

#### 4.1 Residual Plot

Before drawing conclusions from our regression model, it is important to verify that the underlying assumptions are met. To assess linearity and homoscedasticity, we examine the residuals versus fitted values plot.

```{r, echo=FALSE, fig.align='center', fig.height=6}
fitted_values <- final_model$fitted.values
residuals <- final_model$residuals
df_residuals <- data.frame(r = residuals, y_hat = fitted_values)

df_residuals %>%
  ggplot(aes(x = y_hat, y = r)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue") +
  labs(x = "Fitted Values",
       y = "Residuals",
       title = "Residuals Plot of Ladder Score") +
  theme_minimal(base_size = 20)
```

-   **Interpretation**

## [4. Model Diagnostics]{style="font-size:0.8em"}

#### 4.2 Normal Q-Q Plot

To further evaluate the assumptions of our regression model, we examine a normal Q-Q plot of the residuals to check whether the residuals are approximately normally distributed.

```{r, echo=FALSE, fig.align='center'}
std_residuals <- rstandard(final_model)

df_std_residuals <- data.frame(std_r = std_residuals)

df_std_residuals %>%
  ggplot(aes(sample = std_r)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(x = "Theoretical Quantile", 
       y = "Sample Quantile",
       title = "Q-Q Plot of Standardized Residuals") +
  theme_minimal(base_size = 15)
```

-   **Interpretation**

## [4. Model Diagnostics]{style="font-size:0.8em"}

#### 4.3 **Multicollinearity**

We also need to check if there exists any multicollinearity between covariates. We will do this using a correlation table.

::: {.model-table style="border:2px solid #ccc; padding:0em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
col1 <- c("Logged.GDP.per.capita",
          "Social.support",
          "Freedom.to.make.life.choices",
          "Perceptions.of.corruption")

cor_mat <- cor(whr23[, col1])

short_names <- c("Log GDP", "Social Support", "Freedom", "Corruption")

cor_mat <- cor(whr23[, col1])
colnames(cor_mat) <- short_names
rownames(cor_mat) <- short_names

knitr::kable(round(cor_mat, 2), caption = "Correlation Table")
```
:::

**Interpretation:** **Logged GDP per capita** and **Social support** exhibits a relatively high correlation (0.74). This relationship is expected, as economically stronger countries often have greater capacity to provide social support. Additionally, **Social support** encompasses broader factors beyond economic wealth, suggesting that the two variables capture sufficiently distinct aspects of national well-being.

## [5. Interpretation of Coefficients]{style="font-size:0.8em"}

After finalizing our regression model, we interpret the estimated coefficients to understand the specific impact of each predictor on the Ladder Score.

-   **Intercept (-1.444)**

-   **Logged.GDP.per.capita (0.258)**

-   **Social.support (4.141)**

-   **Freedom.to.make.life.choices (2.368)**

-   **Perceptions.of.corruption (-0.869)**

-   $R^2$ (0.8251)

## [6. PCA and Clustering]{style="font-size:0.8em"}

#### 6.1 Proportion of Variance Explained

To better understand the underlying structure of the data and reduce dimensionality, we perform a Principal Component Analysis (PCA) on the selected predictors.

:::: panel-tabset
### PCA Summary Table

```{r, echo=FALSE}
# 1. Prepare and clean the data
col <- c("Ladder.score",
         "Logged.GDP.per.capita",
         "Social.support",
         "Freedom.to.make.life.choices",
         "Perceptions.of.corruption",
         "Country.name")

X <- na.omit(whr23[, col])
countries <- X$Country.name
X_scaled <- scale(as.matrix(X[, 1:5]))

# 2. PCA
PCA <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
PC2 <- PCA$rotation[, 1:2]
X_pca_2d <- X_scaled %*% PC2

# 3. K-means clustering (e.g., 3 clusters)
set.seed(42)
kmeans_res <- kmeans(X_pca_2d, centers = 3, nstart = 25)

# 4. Prepare data frame for plotting
pca_df <- as.data.frame(X_pca_2d)
colnames(pca_df) <- c("PC1", "PC2")
pca_df$Cluster <- as.factor(kmeans_res$cluster)
pca_df$Country <- countries
```

::: {.model-table style="border:2px solid #ccc; padding:0em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
# Get standard deviation and variance explained
std_dev <- PCA$sdev
var_explained <- (std_dev^2) / sum(std_dev^2)
cum_var_explained <- cumsum(var_explained)

# Create a summary data frame
pca_summary_df <- data.frame(
  PC = paste0("PC", 1:length(std_dev)),
  Std_Dev = round(std_dev, 3),
  Proportion_Variance = round(var_explained, 3),
  Cumulative_Variance = round(cum_var_explained, 3)
)

# Display with kable
knitr::kable(pca_summary_df, caption = "PCA Summary Table")
```
:::

### Interpretation

The first principal component (PC1) accounts for approximately **65.9%** of the total variance in the data.

The second component (PC2) contributes an additional **15.6%**, bringing the **cumulative variance explained by the first two components to 81.5%**.
::::

## [6. PCA and Clustering]{style="font-size:0.8em"}

#### 6.1 Proportion of Variance Explained

We can also create a scree plot to better visualize this trend:

```{r, echo=FALSE, fig.align='center'}
pc_df <- data.frame(PC = paste0("PC", 1:length(var_explained)),
                    Variance = var_explained)

pc_df$PC_num <- seq_along(pc_df$Variance)

ggplot(pc_df, aes(x = PC_num, y = Variance)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line(color = "steelblue") +
  scale_x_continuous(breaks = pc_df$PC_num, labels = pc_df$PC) +
  labs(title = "Scree Plot of PCA",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal(base_size = 15)
```

The scree plot shows an obvious elbow pattern in explained variance at the second principal component (PC2)

This affirms our previous conclusion that **PC1 and PC2 capture the majority of the variance** in the data.

## [6. PCA and Clustering]{style="font-size:0.8em"}

#### 6.2 Component Loadings

::: {.model-table style="border:2px solid #ccc; padding:0em; border-radius:6px; background-color:#f9f9f9;"}
```{r, echo=FALSE}
PCA$rotation[, 1:2] %>% pander(caption = "Table of Component Loadings")
```
:::

-   [**note**: PC2 shown here is the negative of that in the report, which does not affect the overall trend and result.]{style="font-size:0.7em"}

## [6. PCA and Clustering]{style="font-size:0.8em"}

#### 6.3 K-means Clustering

To uncover natural groupings among countries based on happiness-related factors, we apply k-means clustering to the 2D principal component projections.

We begin by examining the 2D PCA plot to help determine an appropriate number of clusters.

```{r, echo=FALSE, fig.align='center'}
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 2) +
  labs(title = "2D PCA of WHR23",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal(base_size = 20)
```

## [6. PCA and Clustering]{style="font-size:0.8em"}

#### 6.3 K-means Clustering

```{r, echo=FALSE, fig.align='center'}
selected_countries <- c("Finland", 
                        "New Zealand",
                        "Luxembourg",
                        "United States", 
                        "China",
                        "Japan",
                        "Afghanistan",
                        "Lebanon",
                        "Sierra Leone"
                        )

cluster_colors <- c(
  "1" = "#FF8C00",  # dark orange
  "2" = "#1F78B4",  # steel blue
  "3" = "#E31A1C"   # strong red
)  

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2) +
  geom_text(data = subset(pca_df, Country %in% selected_countries),
            aes(label = Country),
            vjust = -1, size = 3.5, color = "black") +
  labs(title = "2D PCA of WHR23 with K-means Clustering",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal(base_size = 15) +
  scale_color_manual(values = cluster_colors)
```

## Conclusion

Our analysis of the 2023 World Happiness Report identified key factors associated with national happiness. Social support emerged as the strongest positive predictor, followed by freedom to make life choices and GDP per capita, while perceived corruption had a significant negative effect. The final regression model, which included only statistically significant predictors, explained over 82% of the variation in Ladder Scores. To further examine structural patterns in the data, we applied Principal Component Analysis (PCA) and K-means clustering, revealing meaningful groupings of countries by happiness level. These findings highlight the importance of not only economic development but also social support and institutional trust in promoting national well-being. **Rather than focusing solely on GDP, governments should prioritize strengthening social networks and fostering public trust to more effectively enhance happiness.**
